Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
minor edit to appendix

\subsection{Lower bound on likelihood function}\label{ap:MStep}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\theta$. We want to maximise the likelihood function which is difficult to maximise directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\theta)=\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)p(\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\theta)\le \ln~p(\mathbf Y\mid\boldsymbol\theta)
	\end{equation}

>>>>

\subsection{Lower bound on likelihood function}\label{ap:MStep}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. We want to maximise the likelihood function which is difficult to perform directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\theta)=\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)p(\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\theta)\le \ln~p(\mathbf Y\mid\boldsymbol\theta)
	\end{equation}
